Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions

### 2021-02-18 Latex/paper/paper.tex
Adding latex files from appendix & paper

% Paragraph 4: Issues with the data/methods
To date, many differential peak callers (DPCs) have been proposed \citep{song2011identifying,stark2011diffbind,shen2013diffreps,chen2015novel,lun2015csaw,allhoff2016differential}. However, several challenges affect their ability to accurately detect regions of differential activity from the wide range of ChIP-seq experiments (Section \ref{sec:background}). First, differential regions may be both short or broad in length, causing difficulty for methods optimized for a particular type of signal profile \citep{stark2011diffbind,chen2015novel}. Second, methods that pool experimental replicates together \citep{song2011identifying} often exhibit more false positive calls compared to methods that jointly model replicates from each condition \citep{steinhauser2016comprehensive}. Third, the analysis of ChIP-seq data is often subject to complex biases that may vary across the genome, as differences in local read enrichment may depend on quantities such as the total read abundance in a given region or GC content \citep{teng2017accounting}. DPCs that disregard such complex effects or that solely rely on sample-specific global scaling factors or control subtraction methods \citep{stark2011diffbind,shen2013diffreps,chen2015novel,allhoff2016differential} may be prone to detecting spurious differences due to the lack of non-linear normalization methods \citep{lun2015csaw}. Reflecting these limitations, a recent comparison of DPCs demonstrated that current methods tend to detect either a large number of short peaks (low sensitivity) or exhibit a high number of false positive calls (low specificity) in ChIP-seq experiments with broad regions of enrichment \citep{steinhauser2016comprehensive}. Moreover, few methods are able to simultaneously test for differential activity across three or more conditions \citep{chen2015novel,lun2015csaw}, or can classify specific differential combinatorial patterns. Altogether, these limitations can impact the drawing of accurate insights from modern epigenomic studies.

% Paragraph 4: What we propose? Broad peaks, mixture model, and multiple conditions
Here, we propose an efficient and flexible statistical method to identify differential regions of enrichment from epigenomic experiments with diverse signal profiles and collected under common multi-replicate, multi-condition settings. Our method overcomes the limitations of current DPCs with three major features. First, it uses a hidden Markov model (HMM) to account for the diversity in differential enrichment profiles that may result from short and broad epigenomic ChIP-seq data sets. Second, it captures specific differential combinatorial patterns through a novel finite mixture model emission distribution within the HMM's differential state. Each mixture component pertains to a particular differential combinatorial pattern that is formed by the presence or absence of local enrichment across conditions, where a generalized linear model (GLM) is used to model the specific differential combinatorial pattern while accounting for sample- and window-specific normalization factors via offsets. Third, it enables the simultaneous detection and classification of epigenomic changes under three or more conditions, a novelty not yet available in any other DPC algorithm. The presented method offers additional benefits over current HMM-based DPC algorithms \citep{song2011identifying,allhoff2016differential} that include a GLM-based framework with an embedded mixture model, which allows the modeling of covariates of interest as well as the inclusion of model offsets for non-linear normalization, and a fast and accurate parameter estimation scheme via rejection-controlled EM algorithm (RCEM).

@@

Using ChIP-seq data pertaining to histone modifications H3K27me3 and H3K36me3, and to EZH2 from the ENCODE Consortium, we find that current DPCs have difficulty in accurately detecting broad regions of differential enrichment between several common cell lines (Figure~\ref{fig:fig_motivation}). In line with previous findings \citep{steinhauser2016comprehensive}, we observe that even current DPCs designed for broad data \citep{song2011identifying,allhoff2016differential} tend to detect either overly fragmented differential peaks or call regions exhibiting no difference in experimental signal between conditions as differential (Figure~\ref{fig:fig_motivation}A). The low specificity and sensitivity of such methods may impair the biological interpretation of the resulting peak calls in downstream analyses. Methods that rely on candidate peaks may also exhibit a compromised performance due to the limitations of single-sample peak callers in broad data \citep{stark2011diffbind,chen2015novel}. In addition, most current DPCs restrict their application to the analysis of two experimental conditions. For methods that are tailored for the analysis of three or more conditions, the classification of specific differential combinatorial patterns across conditions (or across various epigenomic processes) is still an open problem. The classification of such patterns would allow researchers to, for example, quantify treatment responses on the epigenomic level \citep{clouaire2014cfp1}, or identify sets of processes working together to regulate local chromatin state. We find that the performance of such methods exhibit low sensitivity and specificity in calling differential regions in broad marks (Figure~\ref{fig:fig_motivation}B).
\begin{figure}
\begin{center}
\includegraphics[scale = 0.75]{../../Main/Figure_epigraHMM/Figure_epigraHMM.pdf}
\end{center}
\caption{At the top, differential peak calls from current methods (under false discovery rate control of 0.05) for the histone modification H3K27me3 among cell lines Helas3, Hepg2, and Huvec. Shaded regions indicate  observed differential enrichment, and each vertical line type bordering each region represents a different combinatorial pattern of enrichment across cell lines. Optimal methods would call broad peaks inside shaded regions and no peaks outside them. At the bottom, general overview of the presented method for simultaneous detection and classification of broad and short differential regions of enrichment. This figure appears in color in the electronic version of this article.
\label{fig:fig_motivation}}
\end{figure}

@@

\subsection{Statistical Model}
\label{subsec:hmmmixt}
Let $Y_{hij}$ denote the random variable pertaining to the ChIP read count for genomic window $j$ from sample $i$ of condition $h$, where $j=1,\ldots,M$, $i=1,\ldots,n_{h}$, $h=1,\ldots,G$, and let $y_{hij}$ be the observed count. Here, $n_h$ is the number of samples in condition $h$ and $N = \sum_{h = 1}^G n_h$ is the total number of samples across the $G$ conditions.  At the $j^{th}$ window, let $\mb{y}_{..j}=(y_{11j},\ldots,y_{Gn_{G}j})^\prime$ denote the $N \times 1$ vector of ChIP window read counts across all samples and conditions, and let $\mb{y}=(\mb{y}_{..1},\ldots,\mb{y}_{..M})$ denote the corresponding $N \times M$ matrix of window read counts spanning all windows, samples, and conditions. We assume that each window belongs to one of three possible hidden states: consensus background (state 1), differential (state 2), and consensus enrichment (state 3). Windows exhibiting low (high) enrichment across all conditions will be modeled by an emission distribution pertaining to the consensus background (enrichment) state. Windows exhibiting enrichment under at least one condition, but not all conditions, will be modeled by an emission distribution pertaining to the differential state. If $G$ conditions are of interest, there are $L = 2^G-2$ possible differential combinatorial patterns of enrichment and background across conditions at a given window. The emission distribution pertaining to the differential state models all $L$ possible differential combinatorial patterns via a mixture model with mixture proportions $\bs{\delta} = (\delta_{1},\ldots,\delta_{L})^\prime$, such that $\sum_{l=1}^{L}\delta_{l}=1$ (Web Figure 1).

@@

We assume that read counts pertaining to genomic windows from the consensus background ($r=1$) and consensus enrichment  ($r=3$) states follow a Negative Binomial (NB) distribution with state-specific parameters $\bs{\psi}_{r}=(\mu_{(r,hij)},\phi_{r})^\prime$, with mean $\mu_{(r,hij)}$ and variance $\mu_{(r,hij)}(1+\mu_{(r,hij)}/\phi_{r})$. Assuming independence of read counts across experiments and samples, conditional upon the HMM state, the emission distribution of the consensus background and consensus enrichment states, respectively, can be written as
\begin{align}\label{eq:nondiffdensity}
f_{r}(\mb{y}_{..j}|\bs{\psi}_{r}) &= \prod^{G}_{h=1}\prod^{n_{h}}_{i=1}\frac{\Gamma(y_{hij}+\phi_{r})}{y_{hij}!\Gamma(\phi_{r})}\left(\frac{\phi_{r}}{\mu_{(r,hij)}+\phi_{r}}\right)^{\phi_{r}}\left(\frac{\mu_{(r,hij)}}{\mu_{(r,hij)}+\phi_{r}}\right)^{y_{hij}},\quad r\in\{1,3\},
\end{align}
with $y_{hij}\in\{0,1,2,\ldots\}$, such that $\log(\mu_{(1,hij)})=\beta_{1} + u_{hij}$, $\log(\phi_{1})=\lambda_{1}$, $\log(\mu_{(3,hij)})=\beta_{1} + \beta_{3} + u_{hij}$, and $\log(\phi_{3})=\lambda_{1} + \lambda_{3}$. The offset $u_{hij}$ adjusts for technical artifacts and allows the non-linear normalization of the signal profile across genomic windows, conditions, and samples (Web Appendix B2). When $u_{hij}=0$, $\beta_{1}$ and $\lambda_{1}$ represent the log-mean and log-dispersion, respectively, of read counts pertaining to consensus background state windows, whereas $\beta_{3}$ and $\lambda_{3}$ represent the difference in log-mean and log-dispersion of read counts from consensus enrichment state windows relative to consensus background state windows.

For windows belonging to the differential state ($r = 2$), we assume that the corresponding read counts are modeled by a $L$-component finite mixture model with mixture components that follow a Negative Binomial distribution, where each component corresponds to a particular differential combinatorial pattern. To define these patterns, let us consider the sets $S_{1},\ldots,S_{L}$ that delineate the subset of the $G$ conditions that are enriched in each of the $L$ differential combinatorial patterns. For instance, if $G=3$, the sets $S_{1} =\{1\}$, $S_{2} =\{2\}$, $S_{3} =\{3\}$, $S_{4} =\{1,2\}$, $S_{5} =\{1,3\}$, and $S_{6} =\{2,3\}$ define the six possible differential combinatorial patterns of enrichment and background across three conditions. That is, the set  $S_{1}$ denotes enrichment in only the first condition and background in all others, whereas the set $S_{6}$ denotes enrichment in conditions 2 and 3 and background in condition 1. The presence or absence of enrichment in each of the $L$ sets is encoded into each column of $\mb{x}=(\mb{x}_{1},\ldots,\mb{x}_{L})$, such that $\mb{x}_{l} = (x_{1l},\ldots,x_{Gl})'$, and $x_{hl} = I(h\in S_{l})$ for $l=1,\ldots,L$ and $h = 1,\ldots,G$. That is, $\mb{x}_{l}$ is the $G \times 1$ vector of binary indicator variables denoting which subset of conditions are enriched in pattern (mixture component) $l$. A graphical illustration of our proposed model is provided in Web Figure 1.

@@

\subsubsection{Simulation Results}
Table~\ref{Table:table_simulation} shows the true values and the average relative bias of parameter estimates (and the $2.5^{th}$, $97.5^{th}$ percentiles) from a hundred simulated data sets relative to the scenario of H3K27me3 with $10^5$ genomic windows (see Web Appendix C1 for additional details). Results are shown for different levels of SNR, number of conditions, and number of replicates per condition. Overall, no significant differences regarding the relative bias of parameter estimates were observed across simulations under different number of genomic windows. Depending on the number of conditions, the observed relative bias and the range of the reported percentiles tended to decrease as more replicates were included in the analyses. This effect was particularly significant in scenarios with four conditions with respect to parameters $\beta_{3}$ and $\lambda_{3}$. In general, scenarios with higher SNR showed lower relative bias and variability of the parameter estimates in comparison to scenarios with lower SNR, regardless of the number of conditions or replicates per condition. In scenarios with lower SNR levels or higher number of conditions, these results also highlight the importance of experimental replicates to achieve precise parameter estimates. The proposed estimation approach via EM algorithm led to precise parameter estimates and was robust to a data generating mechanism that was different than the one assumed by the proposed model.
\begin{center}
\input{../../Main/Table_Simulation_ReadCounts/Table_Simulation_ReadCounts.tex}
\end{center}

@@

We compared the genome-wide performance of the presented model with the current DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR. For the methods that require a set of candidate regions to be specified \textit{a priori}, ChIPComp and DiffBind, peaks were called in advance using MACS2 (\cite{zhang2008model}; available at \url{https://github.com/macs3-project/MACS}) and used as input to the software for differential call. We benchmarked methods regarding the coverage of differentially transcribed gene bodies, the number and average size of differential peak calls, $\log_{2}$ fold change (LFC) of read counts, Spearman correlation of $\log_{2}$-transformed read counts between cell lines, and computation time. Metrics for sensitivity and specificity were defined on the window level and based on the coverage of differentially transcribed gene bodies by called peaks (Web Appendix D1). For broad marks, read counts were computed using non-overlapping windows of 500bp. For the remaining short marks, we computed read counts using non-overlapping windows of 250bp. Results presented in this section pertain to the analysis of two cell lines, namely Helas3 and Hepg2. A discussion about the choice of the window size is presented in the Web Appendix D2. Results from the analysis of more than two cell lines are presented in the Web Appendix D3. Data accessing code, data pre-processing steps, method-specific parameters, and code to replicate the presented results are detailed in the Web Appendix A.

@@

In Figure~\ref{fig:fig_broad}A, we show receiver operating characteristic (ROC) curves for various methods and different nominal levels of FDR threshold for differential peaks of H3K36me3. Similar to the analysis of broad histone modification marks presented by \citet{xing2012genome}, we computed the observed true positive rates (false positive rates) on the window-level as the proportion of windows called as differential out of the total number of windows associated (not associated) with differentially transcribed genes (Web Appendix D1). Our method had the best overall performance among all DPCs as its differential peaks were able to cover most of differentially transcribed gene bodies while still maintaining a low number of false positives. Methods that tended to call short peaks, such as ChIPComp and DiffBind, were the ones with the lowest sensitivity among all methods. ChIPComp and DiffBind have been previously shown to be dependent on the set of candidate peaks and to perform best in scenarios with short peaks (Figure~\ref{fig:fig_broad}B; \citet{steinhauser2016comprehensive,lun2015csaw}). In Figure~\ref{fig:fig_broad}C, we show the observed sensitivity (y-axis) and the average differential peak size (kbp; x-axis) for various methods under different nominal FDR levels (the observed FDR is annotated next to each data point). Our model and RSEG, two HMM-based methods, tended to call broader differential peaks and exhibited better sensitivity than other methods. Yet, differential peaks called by RSEG often did not correspond to differential regions of enrichment (Figure~\ref{fig:fig_broad}D), a behavior that has been noted by others \citep{starmer2016detecting} and also seen in simulated data (Figure~\ref{fig:fig_simulation_csaw}). Our HMM-based method with a non-linear normalization scheme via model offsets allowed us to maintain a low observed FDR and a higher sensitivity than other DPCs. In Figure~\ref{fig:fig_broad}F, we show examples of differential peak calls for EZH2. Our method was among the fastest algorithms due to our computational scheme, taking approximately 1 hour to analyze genome-wide data (Figure~\ref{fig:fig_broad}E).
\begin{figure}
\begin{center}
\includegraphics[scale = 0.65]{../../Main/Figure_Broad/Figure_Broad.pdf}
\end{center}
\caption{Analysis of broad ENCODE data. (A): ROC curves of H3K36me3 differential peak calls. (C): sensitivity (y-axis) and average H3K36me3 differential peak size (kbp; x-axis) of various methods under different nominal FDR thresholds (observed FDR annotated next to data points). (B), (D), and (F): example of peak calls from H3K36me3, H3K27me3, and EZH2, respectively, under a nominal FDR control of 0.05. Posterior probabilities of the HMM differential state are shown at the bottom of each panel. (E): computing time of genome-wide analysis from various methods. This figure appears in color in the electronic version of this article.
\label{fig:fig_broad}}
\end{figure}

\subsection{Analysis of ChIP-seq Data From Short Marks}
\label{sec:sharp}
We further evaluated the performance of the proposed method on data sets characterized by short peaks, namely the histone modifications H3K4me3 and H3K27ac and the transcription factor CTCF. The goal of our analysis was to assess whether our method was robust to different types of data and still able to call short differential regions of enrichment. In these scenarios, differential peaks are usually observed in isolated genomic regions and exhibit a high SNR. It has been shown that certain HMM-based approaches, including RSEG, have low accuracy under short histone modification marks and TF data \citep{hocking2016optimizing}. As we show, the model proposed in this article performs comparably to the evaluated DPCs known to perform best in short data (ChIPComp, DiffBind; \cite{steinhauser2016comprehensive}) and appeared to be more efficient regarding the computation time in certain scenarios.

>>>>

% Paragraph 4: Issues with the data/methods
To date, many differential peak callers (DPCs) have been proposed \citep{song2011identifying,stark2011diffbind,shen2013diffreps,chen2015novel,lun2015csaw,allhoff2016differential}. However, several challenges affect their ability to accurately detect regions of differential activity from the wide range of ChIP-seq experiments (Section \ref{sec:background}). First, differential regions may be both short or broad in length, causing difficulty for methods optimized for a particular type of signal profile \citep{stark2011diffbind,chen2015novel}. Second, methods that pool experimental replicates together \citep{song2011identifying} often exhibit more false positive calls compared to methods that jointly model replicates from each condition \citep{steinhauser2016comprehensive}. Third, the analysis of ChIP-seq data is often subject to complex biases that may vary across the genome, as differences in local read enrichment may depend on quantities such as the total read abundance in a given region or GC content \citep{teng2017accounting}. DPCs that disregard such complex effects or that solely rely on sample-specific global scaling factors or control subtraction methods \citep{shen2013diffreps,chen2015novel,allhoff2016differential} may be prone to detecting spurious differences due to the lack of non-linear normalization methods \citep{lun2015csaw}. Reflecting these limitations, a recent comparison of DPCs demonstrated that current methods tend to detect either a large number of short peaks (low sensitivity) or exhibit a high number of false positive calls (low specificity) in ChIP-seq experiments with broad regions of enrichment \citep{steinhauser2016comprehensive}. Moreover, only few methods are able to simultaneously test for differential activity across three or more conditions \citep{stark2011diffbind,chen2015novel,lun2015csaw}, or can classify specific differential combinatorial patterns. Altogether, these limitations can impact the drawing of accurate insights from modern epigenomic studies.

% Paragraph 4: What we propose? Broad peaks, mixture model, and multiple conditions
Here, we propose an efficient and flexible statistical method to identify differential regions of enrichment from epigenomic experiments with diverse signal profiles and collected under common multi-replicate, multi-condition settings. Our method overcomes the limitations of current DPCs with three major features. First, it uses a hidden Markov model (HMM) to account for the diversity in differential enrichment profiles that may result from short and broad epigenomic ChIP-seq data sets. Second, it captures specific differential combinatorial patterns through a novel finite mixture model emission distribution within the HMM's differential state. Each mixture component pertains to a particular differential combinatorial pattern that is formed by the presence or absence of local enrichment across conditions, where a generalized linear model (GLM) is used to model the specific differential combinatorial pattern while accounting for sample- and window-specific normalization factors via offsets. Third, it enables the simultaneous detection and classification of epigenomic changes under three or more conditions, a novelty not yet available in any other DPC algorithm. The presented method, epigraHMM (Figure~\ref{fig:fig_motivation}), offers additional benefits over current HMM-based DPC algorithms \citep{song2011identifying,allhoff2016differential} that includes a flexible GLM-based framework with an embedded mixture model, which allows the modeling of covariates of interest, the inclusion of model offsets for non-linear normalization and GC-bias adjustment, and a fast and accurate parameter estimation scheme via rejection-controlled EM algorithm (RCEM).

@@

Using ChIP-seq data pertaining to histone modifications H3K27me3 and H3K36me3, and to EZH2 from the ENCODE Consortium, we find that current DPCs have difficulty in accurately detecting broad regions of differential enrichment between several common cell lines (Figure~\ref{fig:fig_motivation}). In line with previous findings \citep{steinhauser2016comprehensive}, we observe that even current DPCs designed for broad data \citep{song2011identifying,allhoff2016differential} tend to detect either overly fragmented differential peaks or call regions exhibiting no difference in experimental signal between conditions as differential (Sections \ref{sec:sim2} and \ref{sec:broad}). The low specificity and sensitivity of such methods may impair the biological interpretation of the resulting peak calls in downstream analyses. Methods that rely on candidate peaks may also exhibit a compromised performance due to the limitations of single-sample peak callers in broad data \citep{stark2011diffbind,chen2015novel}. In addition, most current DPCs restrict their application to the analysis of two experimental conditions. For methods that are tailored for the analysis of three or more conditions, the classification of specific differential combinatorial patterns across conditions (or across various epigenomic processes) is still an open problem. The classification of such patterns would allow researchers to, for example, quantify treatment responses on the epigenomic level \citep{clouaire2014cfp1}, or identify sets of processes working together to regulate local chromatin state. We find that the performance of such methods exhibit low sensitivity and specificity in calling differential regions in broad marks (Figure~\ref{fig:fig_motivation}).
\begin{figure}
\begin{center}
\includegraphics[scale = 0.75]{../../Main/Figure_epigraHMM/Figure_epigraHMM.pdf}
\end{center}
\caption{At the top, differential peak calls from current methods (under false discovery rate control of 0.05) for the histone modification H3K27me3 among cell lines Helas3, Hepg2, and Huvec. Shaded regions indicate  observed differential enrichment, and each vertical line type bordering each region represents a different combinatorial pattern of enrichment across cell lines. Optimal methods would call broad peaks inside shaded regions and no peaks outside them. At the bottom, general overview of the presented method for simultaneous detection and classification of broad and short differential regions of enrichment. This figure appears in color in the electronic version of this article.
\label{fig:fig_motivation}}
\end{figure}

@@

\subsection{Statistical Model}
\label{subsec:hmmmixt}
Let $Y_{hij}$ denote the random variable pertaining to the ChIP read count for genomic window $j$ from sample $i$ of condition $h$, where $j=1,\ldots,M$, $i=1,\ldots,n_{h}$, $h=1,\ldots,G$, and let $y_{hij}$ be the observed count. Here, $n_h$ is the number of samples in condition $h$ and $N = \sum_{h = 1}^G n_h$ is the total number of samples across the $G$ conditions.  At the $j^{th}$ window, let $\mb{y}_{..j}=(y_{11j},\ldots,y_{Gn_{G}j})^\prime$ denote the $N \times 1$ vector of ChIP window read counts across all samples and conditions, and let $\mb{y}=(\mb{y}_{..1},\ldots,\mb{y}_{..M})$ denote the corresponding $N \times M$ matrix of window read counts spanning all windows, samples, and conditions. We assume that each window belongs to one of three possible hidden states: consensus background (state 1), differential (state 2), and consensus enrichment (state 3). Windows exhibiting low (high) enrichment across all conditions will be modeled by an emission distribution pertaining to the consensus background (enrichment) state. Windows exhibiting enrichment under at least one condition, but not all conditions, will be modeled by an emission distribution pertaining to the differential state. If $G$ conditions are of interest, there are $L = 2^G-2$ possible differential combinatorial patterns of enrichment and background across conditions at a given window. The emission distribution pertaining to the differential state models all $L$ possible differential combinatorial patterns via a mixture model with mixture proportions $\bs{\delta} = (\delta_{1},\ldots,\delta_{L})^\prime$, such that $\sum_{l=1}^{L}\delta_{l}=1$.

@@

We assume that read counts pertaining to genomic windows from the consensus background ($r=1$) and consensus enrichment  ($r=3$) states follow a Negative Binomial (NB) distribution with state-specific parameters $\bs{\psi}_{r}=(\mu_{(r,hij)},\phi_{r})^\prime$, with mean $\mu_{(r,hij)}$ and variance $\mu_{(r,hij)}(1+\mu_{(r,hij)}/\phi_{r})$. Assuming independence of read counts across experiments and samples, conditional upon the HMM state, the emission distribution of the consensus background and consensus enrichment states, respectively, can be written as
\begin{align}\label{eq:nondiffdensity}
f_{r}(\mb{y}_{..j}|\bs{\psi}_{r}) &= \prod^{G}_{h=1}\prod^{n_{h}}_{i=1}\frac{\Gamma(y_{hij}+\phi_{r})}{y_{hij}!\Gamma(\phi_{r})}\left(\frac{\phi_{r}}{\mu_{(r,hij)}+\phi_{r}}\right)^{\phi_{r}}\left(\frac{\mu_{(r,hij)}}{\mu_{(r,hij)}+\phi_{r}}\right)^{y_{hij}},\quad r\in\{1,3\},
\end{align}
with $y_{hij}\in\{0,1,2,\ldots\}$, such that $\log(\mu_{(1,hij)})=\beta_{1} + u_{hij}$, $\log(\phi_{1})=\lambda_{1}$, $\log(\mu_{(3,hij)})=\beta_{1} + \beta_{3} + u_{hij}$, and $\log(\phi_{3})=\lambda_{1} + \lambda_{3}$. The offset $u_{hij}$ adjusts for technical or biological artifacts (such as GC-content bias) and allows the non-linear normalization of the signal profile across genomic windows, conditions, and samples (Web Appendix B2). When $u_{hij}=0$, $\beta_{1}$ and $\lambda_{1}$ represent the log-mean and log-dispersion, respectively, of read counts pertaining to consensus background state windows, whereas $\beta_{3}$ and $\lambda_{3}$ represent the difference in log-mean and log-dispersion of read counts from consensus enrichment state windows relative to consensus background state windows.

For windows belonging to the differential state ($r = 2$), we assume that the corresponding read counts are modeled by a $L$-component finite mixture model with mixture components that follow a Negative Binomial distribution, where each component corresponds to a particular differential combinatorial pattern. To define these patterns, let us consider the sets $S_{1},\ldots,S_{L}$ that delineate the subset of the $G$ conditions that are enriched in each of the $L$ differential combinatorial patterns. For instance, if $G=3$, the sets $S_{1} =\{1\}$, $S_{2} =\{2\}$, $S_{3} =\{3\}$, $S_{4} =\{1,2\}$, $S_{5} =\{1,3\}$, and $S_{6} =\{2,3\}$ define the six possible differential combinatorial patterns of enrichment and background across three conditions. That is, the set  $S_{1}$ denotes enrichment in only the first condition and background in all others, whereas the set $S_{6}$ denotes enrichment in conditions 2 and 3 and background in condition 1. The presence or absence of enrichment in each of the $L$ sets is encoded into each column of $\mb{x}=(\mb{x}_{1},\ldots,\mb{x}_{L})$, such that $\mb{x}_{l} = (x_{1l},\ldots,x_{Gl})'$, and $x_{hl} = I(h\in S_{l})$ for $l=1,\ldots,L$ and $h = 1,\ldots,G$. That is, $\mb{x}_{l}$ is the $G \times 1$ vector of binary indicator variables denoting which subset of conditions are enriched in pattern (mixture component) $l$.

@@

\subsubsection{Simulation Results}
Table~\ref{Table:table_simulation} shows the true values and the average relative bias of parameter estimates (and the $2.5^{th}$, $97.5^{th}$ percentiles) from a hundred simulated data sets relative to the scenario of H3K27me3 with $10^6$ genomic windows (see Web Appendix C1 for additional details). Results are shown for the SNR level estimated from the ENCODE project (mean count ratio between peak and enrichment regions of 3.21), different number of conditions, and different number of replicates per condition. Overall, no significant differences regarding the relative bias of parameter estimates were observed across simulations under different number of genomic windows. Depending on the number of conditions, the observed relative bias and the range of the reported percentiles tended to decrease as more replicates were included in the analyses. This effect was particularly significant in scenarios with four conditions with respect to parameters $\beta_{3}$ and $\lambda_{3}$. In general, scenarios with higher SNR showed lower relative bias and variability of the parameter estimates in comparison to scenarios with lower SNR, regardless of the number of conditions or replicates per condition. In scenarios with higher number of conditions, these results highlight the importance of experimental replicates to achieve precise parameter estimates. The proposed estimation approach via EM algorithm led to precise parameter estimates and was robust to a data generating mechanism that was different than the one assumed by the proposed model.
\begin{center}
\input{../../Main/Table_Simulation_ReadCounts/Table_Simulation_ReadCounts.tex}
\end{center}

@@

We compared the genome-wide performance of the presented model with the current DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR. For the methods that require a set of candidate regions to be specified \textit{a priori}, ChIPComp and DiffBind, peaks were called in advance using MACS2 (\cite{zhang2008model}; available at \url{https://github.com/macs3-project/MACS}) and used as input to the software for differential call. We benchmarked methods regarding the coverage of differentially transcribed gene bodies, the number and average size of differential peak calls, $\log_{2}$ fold change (LFC) of read counts, Spearman correlation of $\log_{2}$-transformed read counts between cell lines, and computation time. Metrics for sensitivity and specificity were defined on the window level and based on the coverage of differentially transcribed gene bodies by called peaks (Web Appendix D1). For broad marks, read counts were computed using non-overlapping windows of 500bp. For the remaining short marks, we computed read counts using non-overlapping windows of 250bp. Results presented in this section pertain to the analysis of two cell lines, namely Helas3 and Hepg2. Analyses were adjusted by input control samples in methods that are designed to do so (ChIPComp, DiffBind, diffReps, and THOR). We did not observe a significant improvement in performance by accounting for input control effect or GC-content bias with epigraHMM and did not adjust our analyses for these effects. A discussion about the choice of the window size is presented in the Web Appendix D2. Results from the analysis of more than two cell lines are presented in the Web Appendix D3. Data accessing code, data pre-processing steps, method-specific parameters, and code to replicate the presented results are detailed in the Web Appendix A.

@@

In Figure~\ref{fig:fig_broad}A, we show receiver operating characteristic (ROC) curves for various methods and different nominal levels of FDR threshold for differential peaks of H3K36me3. Similar to the analysis of broad histone modification marks presented by \citet{xing2012genome}, we computed the observed true positive rates (false positive rates) on the window-level as the proportion of windows called as differential out of the total number of windows associated (not associated) with differentially transcribed genes (Web Appendix D1). Our method had the best overall performance among all DPCs as its differential peaks were able to cover most of differentially transcribed gene bodies while still maintaining a low number of false positives. Methods that tended to call short peaks, such as ChIPComp and DiffBind, were the ones with the lowest sensitivity among all methods. ChIPComp and DiffBind have been previously shown to be dependent on the set of candidate peaks and to perform best in scenarios with short peaks (Figure~\ref{fig:fig_broad}B; \citet{steinhauser2016comprehensive,lun2015csaw}). In Figure~\ref{fig:fig_broad}C, we show the observed sensitivity (y-axis) and the average differential peak size (kbp; x-axis) for various methods under different nominal FDR levels (the observed FDR is annotated next to each data point). Our model and RSEG, two HMM-based methods, tended to call broader differential peaks and exhibited better sensitivity than other methods. Yet, differential peaks called by RSEG often did not correspond to differential regions of enrichment (Figure~\ref{fig:fig_broad}D), a behavior that has been noted by others \citep{starmer2016detecting} and also seen in simulated data (Figure~\ref{fig:fig_simulation_csaw}). Our HMM-based method with a non-linear normalization scheme via model offsets allowed us to maintain a low observed FDR and a higher sensitivity than other DPCs. In Figure~\ref{fig:fig_broad}F, we show examples of differential peak calls for EZH2. Our method was among the most efficient algorithms due to our computational scheme, taking approximately 30 minutes to analyze genome-wide data (Figure~\ref{fig:fig_broad}E).
\begin{figure}
\begin{center}
\includegraphics[scale = 0.65]{../../Main/Figure_Broad/Figure_Broad.pdf}
\end{center}
\caption{Analysis of broad ENCODE data. (A): ROC curves of H3K36me3 differential peak calls. (C): sensitivity (y-axis) and average H3K36me3 differential peak size (kbp; x-axis) of various methods under different nominal FDR thresholds (observed FDR annotated next to data points). (B), (D), and (F): example of peak calls from H3K36me3, H3K27me3, and EZH2, respectively, under a nominal FDR control of 0.05. Posterior probabilities of the HMM differential state are shown at the bottom of each panel. (E): computing time and peak memory usage of genome-wide analysis from various methods. This figure appears in color in the electronic version of this article.
\label{fig:fig_broad}}
\end{figure}

\subsection{Analysis of ChIP-seq Data From Short Marks}
\label{sec:sharp}
We further evaluated the performance of the proposed method on data sets characterized by short peaks, namely the histone modifications H3K4me3 and H3K27ac and the transcription factor CTCF. The goal of our analysis was to assess whether our method was robust to different types of data and still able to call short differential regions of enrichment. In these scenarios, differential peaks are usually observed in isolated genomic regions and exhibit a high SNR. It has been shown that certain HMM-based approaches, including RSEG, have low accuracy under short histone modification marks and TF data \citep{hocking2016optimizing}. As we show, the model proposed in this article performs comparably to the evaluated DPCs known to perform best in short data (ChIPComp, DiffBind; \cite{steinhauser2016comprehensive}).

We calculated the LFC and the Spearman correlation between cell lines Helas3 and Hepg2 based on ChIP-seq read counts mapped onto differential peaks called by each method. Read counts were previously normalized by the median log-ratios of each replicate over the geometric mean to avoid spurious differences due to sequencing depth. As these marks are characterized by short peaks, ideal methods would show high absolute LFC and negative correlation between read counts mapped on differential peaks. Figure~\ref{fig:fig_sharp} shows the main results from our analysis using short data sets. In Figures~\ref{fig:fig_sharp} A and C we show the median LFC and the Spearman correlation of ChIP-seq counts for differential CTCF and H3K4me3 peak calls (sorted by the absolute LFC), respectively, under a nominal FDR control of 0.05. We present separate curves regarding the signal of observed enrichment to better characterize the direction of change. The results show that the HMM-based methods RSEG and THOR were among those with the lowest absolute LFC among all methods, which confirms their sub optimal performance in the scenario of short peaks \citep{hocking2016optimizing}. In addition, we observed that ChIPComp had the best performance overall as it was able to call differential peaks with the highest absolute LFC and the lowest correlation between read counts of the two analyzed cell lines. Our model was able to properly call truly short differential peaks (Figures~\ref{fig:fig_sharp} B, D, and F) and was comparable to the non-HMM based methods regarding the computation time (Figure~\ref{fig:fig_sharp}E), jointly calling differential peaks in less than 1.5 hour.
\begin{figure}
\begin{center}
\includegraphics[scale = 0.65]{../../Main/Figure_Narrow/Figure_Narrow.pdf}
\end{center}
\caption{Analysis of short ENCODE data. (A) and (C): median LFC and correlation between cell lines of ChIP-seq counts from differential peaks for CTCF and H3K4me3, respectively. (B), (D), and (F): example of peak calls from CTCF, H3K4me3, and H3K27ac, respectively. Posterior probabilities of the HMM differential state are shown at the bottom of each panel. (E): computing time and peak memory usage of genome-wide analysis from various methods. Results are shown under a nominal FDR control of 0.05. This figure appears in color in the electronic version of this article.
\label{fig:fig_sharp}}
\end{figure}
